---
title: Coping with climate change.
subtitle: Implications of the Yellow-Bellied Marmot's (*Marmota flavivientris*) body mass evolution in the last half-century.
author: Augustin Birot (300444988)
date: last-modified
affiliation: Department of Biology
comps-proposal: true # or false
committee:
  - name: Julien Martin
    role: supervisor 
  - name: Vincent Careau
    role: TAC member
  - name: Roslyn Dakin
    role: TAC member
degree: PhD


bibliography: [research_proposal.bib, packages.bib]
csl: american-statistical-association.csl

format:
  #html:
  pdf:
    toc: true
editor: 
  markdown: 
    wrap: 72
---


{{<pagebreak>}}

# Introduction

## Climate change (C.C.)

* The importance of C.C. and its impact in the near future is no longer in doubt, *at the point where it's sad to have to remind people of them once again.*[@intergovernmentalpanelonclimatechangeipccClimateChange20222023]

* Broadly, C.C. is [...] (Polar melting, etc.)

* Which even impact human society (e.g., winter in Ottawa isn't the same anymore: Rideau Canal ice skating future is in jeopardy, the number of days with under -20°C is expected to severly decrease in the near future, etc. (!!! + FACT CHECK everything!))

* Main/Precise impacts of C.C. in natural environments

  * **Raising T°**: Explain + study case (!!!)

  * **Changing season length**: Explain + study case (try to find something at RMBL !!!)

  * **Environmental predictibility**: Explain + study case (!!!)

  * **Drought events**: Explain + study case (!!!)

  * **Extreme weather events**: Explain + study case (!!!)

  * etc.

Ok, so, now, how does C.C. and these precise perturbations impacts concretely natural population? *Study cases* (!!!)


**universal C.C. responses (C.C.)**



## Body size as a Life-History Traits (LHT) and expected effect of global warming

C.C. is expected to impact life history traits [LHT, i.e., traits impacting directly survival and reproduction, so individual's fitness @Roff1992TheEO].

### Link with hibernation
Body mass for hibernating species is so a LHT as it's usually a determining factor for survival over hibernation and reproduction.

Body mass is a LHT as in many species it has direct impact on survival and reproduction (explain + !!!)


[@daufresneGlobalWarmingBenefits2009], [@gardnerDecliningBodySize2011], [@guillemainWinteringFrenchMallard2010], [@sheridanShrinkingBodySize2011], [@yom-tovRecentIncreaseBody2008], [@ozgulCoupledDynamicsBody2010]
[@gienappEvolutionaryDynamicsResponse2014]

And some authors argue that a decreasing body size could be one the universal C.C. response (!!!)

Because Bergmann's rules, which state that smaller body size should be expected in warmer environment as it raises the surface to volume ratio, thus favoring heat dissipation [@bergmanncRelationshipsHeatConservation1847].

Thus, a general shriking body size is expected with global warming.
However opposite results at higher latitude yield objections of this theory and raise the need of more general study about that.
Furthermore, these opposite results (i.e., increasing body mass at higher latitudes) can also been explain as C.C. is synonym to milder conditions in those latitudes. This change Allows individuals to forage more and thus gain mass.


## Q.G. and animal models

Body mass and LHT shifts expected with climate change, **evolution** expected. To test that --> **Animal Models**! [@kruukEstimatingGeneticParameters2004]

[@10.1093/acprof:oso/9780199674237.001.0001] <!-- Q.G. in the wild -->


## LHT coevolution
Traits can't evolve alone Gould & Lewontin (1979)

Need to show that with multivariate animal model, but no one has enough power for the models [@teplitskyEvolutionaryPotentialConstraints2014]

## POLS

[@dammhahnPacelifeSyndromesFramework2018]

## Phenotypic plasticity vs microevolution


## I * E, G * E (individual variation in their plasticity)

[@nusseyEvolutionaryEcologyIndividual2007]

Link with body mass, individual can vary in their growing speed

<!-- I * A and G * A -->

## Species and study site

A wild Yellow-Bellied Marmot (*Marmota flaviventris*, "YBM") population in the Upper East River Valley, Colorado, USA, is the subject of one of the longest-term study in the world (1962 - today).
YBM is a ground-dwelling sciurid (rodentia, sciurideae) inhabiting alpine habitats in western North America with a life cycle divided between an "active season" representing approximately a third of the year (from May to September) where individuals must forage to reach a threshold body mass in order to survive hibernation for the remainder of the time [@armitageMarmotBiologySociality2014].
Individuals experience high seasonal flutuation in body mass, with a critical threshold to be reached before the onset of hibernation in order to 1) survive through the next active season and 2) have sufficient energy left for hibernation [which occurs in the first weeks of the active season, @armitageVernalBehaviourYellowbellied1965; @armitageMarmotBiologySociality2014].
Consequently, body mass is consider being a critical LHT for the marmots.
YBM lives in colonies composed usually by one or more matriline with on adult males, multiple adult females and their offspring [@armitageMarmotBiologySociality2014]. Our population is composed of seven main colonies divided between an "up" and a "down valley" with a elevation difference arround 300m ("up" = 3,000m; "down" = 2,700m) implying some difference in weather [e.g., delayed snowmel and vegetation growth onset, temperature difference up to 2 °C, @blumsteinYellowbelliedMarmotsMarmota2004; @armitageMarmotBiologySociality2014] and so delayed emergence up to two weeks in the up-valley [@blumsteinSOCIALEFFECTSEMERGENCE2009; @monclusYellowbelliedMarmotsNot2014].
This two differents condition offers an amazing opportunity to test the impact on environment on several factors while working in natural conditions.

This hibernation (life) cycle is highly environemntally dependant, with the onset and end of the active season believed to be mediated mostly by weather variable such as temperature and snow cover of the region [@armitageMarmotBiologySociality2014]. Thus, body mass is expected to be a keystone phenotypic trait for the marmots. It is therefore crucial to understand how this trait and this species responds to global warming, both for conservaion purposes and to elucidate links between phenotype and environment.



### Body mass increase in YBM

An important body mass increase has been observed in this population over the past half-century (estimated arround 600 g for the adult females). Precedent studies attributed this major change mostly to phenotypic plasticity [@ozgulCoupledDynamicsBody2010]. This hypothese made in fact a lot of sense, with climate change active season is getting longer (milder condition, higher temperature, less snow, shorter winter, etc.), hence marmots have more time to forage, gain weight, and the hibernation period is getting shorter so less time for the individuals to lose mass, at the end of the day, we have heavier individuals, makes sense!
However, using animal models to properly assess the genetic attributable part of this change, thus estimating explicitely the body mass' evolutionary signal for the adult females over the time cohort (i.e., year of birth) during the study period, we found an increase, at the genetic scale, estimated arround 400 g (Birot & Martin, Manuscript in progress). So, in fact, arround two third of the body mass increase seems to be due to evolution, not just plasticity.
Furthermore, although the lengthening active season is indeed a good potential explanation for the body mass increase through phenotypic plasticity, it doesn't match with the observed evolutionary signal. If the main selective pressure on body mass is survival through hibernation (i.e., heavier individuals having more chance to survive through winter as they have more ressources), then the expected evolutionary response (i.e., average body mass increase) is occuring when the pressure is decreasing, which doesn't makes sense!
Hence, knowing all that, we now need to reconsider the evolutionary scenario behind this major phenotypic change.

I will explore which environmental factors could have triggered this shift, but also the mechanism behind this increase and finally the potential implication for the population's future. <!-- DEVELOPP, but not too much as you developp in the "chapters" -->





{{<pagebreak>}}


# Chapter 1 - Mechanisms
*Marmot's Biology: What mechanisms are behind the body mass increase?*

Growth? Baseline? Both?


Double random (Intercept, Slope)


```{r}
#| echo: false
#| message: false
#| warning: false
#| eval: true
#| label: fig-juvenile-trend
#| fig-cap: Body mass trend over time cohort for females (black) and males (red) juveniles compared between a) the beginning of the active season (birth weight) and b) the end of the season (mass on August 15th).

source("/Users/augustinbirot/Desktop/juvenile_mass.r")

cowplot::plot_grid(plot2_jun, plot2_aug, ncol = 2)
```

{{<pagebreak>}}

# Chapter 2 - Methodology
*Methodology: $I * E$ detection with double random mixed models*

[@nusseyEvolutionaryEcologyIndividual2007] --> double random

So we're doing something different --> examining the residuals of the model (if I * E, still a lot of residual variance ?)


**Look at this one: [@westneatBiologyHiddenResidual2015a]


DHGLM, brms, Julien's code

Vve (Variance dans la variance résiduel, estime la variance résiduel pour chaque individu et regarde la variance dans cette variance résiduelle, si I * E Vve > 0)

Attention aux modèle débalancés si pas d'effet fixes
corrige pour les variations par effet fixes, puis test pour le I * E, si y'en a tu pexu chercher la variable environnemental pour lesquels on a de la variation dans la plasitcité (I * E)

Ned Dotchermann


{{<pagebreak>}}

# Chapter 3 - Triggers
*Marmot's Biology: Which environemntal factors have triggered the phenotypic shift?*

E1 - E10 (T°, Precipitation, ...), Seasonal Gradient

Predators, Diet?

{{<pagebreak>}}

# Chapter 4 - Implications
*Marmot's Biology: What could be the implications of that for the population's future?*

Manuscript models Body Mass/active season with survival
=> Phenological mismatch?? (e.g., thermal stress)


{{<pagebreak>}}

# Significance and impacts

<!--
# Research objectives and hypotheses

Specific hypotheses (in the case of hypothesis-testing science) or
objectives (in the case of descriptive science), and the underlying
logic, for each proposed thesis chapter (minimum three for PhD students)

# Proposed methodologies

Include outlines of data collection and data analysis

Outline any special skills, equipment, or techniques that will be
required.

In case you want to add a wide table that needs to be on a landscape
page you can do like in @tbl-sideway. Have a look at the `qmd` for the
code.

You need to put the table in a pandoc div (delimited by :::) indicating
the side way rotation

It also works for figures just use `sidewaysfigure` instead of
`sidewaystable`

::: {.sidewaystable data-latex=""}
```{r}
#| label: tbl-sideway
#| echo: false
#| tbl-cap: This is a wide table
library(gt)
exibble |>
  gt(
    rowname_col = "row",
    groupname_col = "group",
    locale = "fr"
  ) |>
  fmt_number() |>
  fmt_date(
    columns = date,
    date_style = "yMEd"
  ) |>
  fmt_datetime(
    columns = datetime,
    format = "EEEE, MMMM d, y",
    locale = "en"
  ) |>
  tab_options(table.width = pct(100))

```
:::


# Expected Products

A list of the (usually 1/3) articles expected for completion of the
MSC/PhD. Think of each as a fantasy abstract.

1.  The Evolution of Human Uniqueness. [Bob Hette & Julien
    Martin.]{.underline}

    There is wide agreement that humans are unique. But it is rarely
    asked: Have they always been unique? In this paper, we derive a
    general uniqueness theorem that states that any unique species must
    have been less unique at some point in the past. This theorem has
    implications for the achievement of world peace.

2.  Niche Construction, Disease Transmission, and Global Warming: Chaos
    Deconstructs the Block Chain. [Bob Hette, Igor Tex & Julien
    Martin]{.underline}

    How many buzz words can be combined in a title? In this paper, we
    find that the natural limit on buzz words in any specific scholarly
    field is always given by the triangular number $T_n$ where $n$ is
    one more than the number of tenured faculty in the field.

3.  A Bayesian Hierarchical Survival Analysis of Counting Sheep. [Bob
    Hette & Julien Martin]{.underline}

    Just like all those other models of counting sheep, but this one
    will be Bayesian.

# Scientific value and rationale for the research

# Course selection

not necessarily needed for comprehensive exam proposal

# Schedule

Outline of the calendar for the work. This is necessarily rough and
subject to change. Usually include a Gantt chart

## Completed to date


## Six month plan
-->

{{< pagebreak >}}

# References

::: {refs}
:::
