---
title: Coping with climate change.
subtitle: Implications of the Yellow-Bellied Marmot's (*Marmota flavivientris*) body mass evolution in the last half-century.
author: Augustin Birot (300444988)
date: last-modified
affiliation: Department of Biology
comps-proposal: false # or false
committee:
  - name: Julien Martin
    role: supervisor 
  - name: Vincent Careau
    role: TAC member
  - name: Roslyn Dakin
    role: TAC member
degree: PhD


bibliography: [research_proposal.bib, packages.bib]
csl: american-statistical-association.csl

format:
  #html:
  pdf:
    toc: true
editor: 
  markdown: 
    wrap: 72
---


{{<pagebreak>}}

# Introduction

## Climate change (C.C.)

* The importance of C.C. and its impact in the near future is no longer in doubt [@intergovernmentalpanelonclimatechangeipccClimateChange20222023]

* Broadly, C.C. is [...] (Polar melting, etc.

* Which even impact human society (e.g., winter in Ottawa isn't the same anymore: Rideau Canal ice skating future is in jeopardy, the number of days with under -20°C is expected to severly decrease in the near future, etc. (!!! + FACT CHECK everything!))

* Main/Precise impacts of C.C. in natural environments

  * **Raising T°**: Explain + study case (!!!)

  * **Changing season length**: Explain + study case (try to find something at RMBL !!!)

  * **Environmental predictibility**: Explain + study case (!!!)

  * **Drought events**: Explain + study case (!!!)

  * **Extreme weather events**: Explain + study case (!!!)

  * etc.

Ok, so, now, how does C.C. and these precise perturbations impacts concretely natural population? *Study cases* (!!!)


**universal C.C. responses (C.C.)**



## Body size as a Life-History Traits (LHT) and expected effect of global warming

C.C. is expected to impact life history traits [LHT, i.e., traits impacting directly survival and reproduction, so individual's fitness @Roff1992TheEO].

Body mass is a LHT as in many species it has direct impact on survival and reproduction (explain + !!!)


[@daufresneGlobalWarmingBenefits2009], [@gardnerDecliningBodySize2011], [@guillemainWinteringFrenchMallard2010], [@sheridanShrinkingBodySize2011], [@yom-tovRecentIncreaseBody2008], [@ozgulCoupledDynamicsBody2010]

And some authors argue that a decreasing body size could be one the universal C.C. response (!!!)

Because Bergmann's rules, which state that smaller body size should be expected in warmer environment as it raises the surface to volume ratio, thus favoring heat dissipation [@bergmanncRelationshipsHeatConservation1847].

Thus, a general shriking body size is expected with global warming.
However opposite results at higher latitude yield objections of this theory and raise the need of more general study about that.
Furthermore, these opposite results (i.e., increasing body mass at higher latitudes) can also been explain as C.C. is synonym to milder conditions in those latitudes. This change Allows individuals to forage more and thus gain mass.


## Q.G. and animal models

[@kruukEstimatingGeneticParameters2004]
[@10.1093/acprof:oso/9780199674237.001.0001] <!-- Q.G. in the wild -->


## LHT coevolution
Traits can't evolve alone Gould & Lewontin (1979)

Need to show that with multivariate animal model, but no one has enough power for the models [@teplitskyEvolutionaryPotentialConstraints2014]

## POLS


## Phenotypic plasticity vs microevolution


## I * E, G * E (individual variation in their plasticity)

[@nusseyEvolutionaryEcologyIndividual2007]

Link with body mass, individual can vary in their growing speed

I * A and G * A

## Species and study site

### Phenological mismatch??

<!-- It provides the scientific basis for the study and a (brief) review of
the current literature, the key question(s) for you research leading -up
to the next section about your specific hypotheses or objectives.

This template is developed for Quarto [@quarto]. You can add tables,
figures and code using R [@base], Python, Julia or Observables. For more
info, go to the Quarto website <https://quarto.org/> -->


{{<pagebreak>}}


# Chapter 1 - Mechanisms
*Marmot's Biology: What mechanisms are behind the body mass increase?*

Growth? Baseline? Both?


Double random (Intercept, Slope)


```{r}
#| echo: false
#| message: false
#| warning: false
#| eval: true
#| label: fig-juvenile-trend
#| fig-cap: Body mass trend over time cohort for females (black) and males (red) juveniles compared between a) the beginning of the active season (birth weight) and b) the end of the season (mass on August 15th).

source("/Users/augustinbirot/Desktop/juvenile_mass.r")

cowplot::plot_grid(plot2_jun, plot2_aug, ncol = 2)
```

{{<pagebreak>}}

# Chapter 2 - Methodology
*Methodology: $I * E$ detection with double random mixed models*

[@nusseyEvolutionaryEcologyIndividual2007] --> double random

So we're doing something different --> examining the residuals of the model (if I * E, still a lot of residual variance ?)


**Look at this one: [@westneatBiologyHiddenResidual2015a]

{{<pagebreak>}}

# Chapter 3 - Triggers
*Marmot's Biology: Which environemntal factors have triggered the phenotypic shift?*

E1 - E10 (T°, Precipitation, ...), Seasonal Gradient

Predators, Diet?

{{<pagebreak>}}

# Chapter 4 - Implications
*Marmot's Biology: What could be the implications of that for the population's future?*

Manuscript models Body Mass/active season with survival


<!--
# Research objectives and hypotheses

Specific hypotheses (in the case of hypothesis-testing science) or
objectives (in the case of descriptive science), and the underlying
logic, for each proposed thesis chapter (minimum three for PhD students)

# Proposed methodologies

Include outlines of data collection and data analysis

Outline any special skills, equipment, or techniques that will be
required.

In case you want to add a wide table that needs to be on a landscape
page you can do like in @tbl-sideway. Have a look at the `qmd` for the
code.

You need to put the table in a pandoc div (delimited by :::) indicating
the side way rotation

It also works for figures just use `sidewaysfigure` instead of
`sidewaystable`

::: {.sidewaystable data-latex=""}
```{r}
#| label: tbl-sideway
#| echo: false
#| tbl-cap: This is a wide table
library(gt)
exibble |>
  gt(
    rowname_col = "row",
    groupname_col = "group",
    locale = "fr"
  ) |>
  fmt_number() |>
  fmt_date(
    columns = date,
    date_style = "yMEd"
  ) |>
  fmt_datetime(
    columns = datetime,
    format = "EEEE, MMMM d, y",
    locale = "en"
  ) |>
  tab_options(table.width = pct(100))

```
:::


# Expected Products

A list of the (usually 1/3) articles expected for completion of the
MSC/PhD. Think of each as a fantasy abstract.

1.  The Evolution of Human Uniqueness. [Bob Hette & Julien
    Martin.]{.underline}

    There is wide agreement that humans are unique. But it is rarely
    asked: Have they always been unique? In this paper, we derive a
    general uniqueness theorem that states that any unique species must
    have been less unique at some point in the past. This theorem has
    implications for the achievement of world peace.

2.  Niche Construction, Disease Transmission, and Global Warming: Chaos
    Deconstructs the Block Chain. [Bob Hette, Igor Tex & Julien
    Martin]{.underline}

    How many buzz words can be combined in a title? In this paper, we
    find that the natural limit on buzz words in any specific scholarly
    field is always given by the triangular number $T_n$ where $n$ is
    one more than the number of tenured faculty in the field.

3.  A Bayesian Hierarchical Survival Analysis of Counting Sheep. [Bob
    Hette & Julien Martin]{.underline}

    Just like all those other models of counting sheep, but this one
    will be Bayesian.

# Scientific value and rationale for the research

# Course selection

not necessarily needed for comprehensive exam proposal

# Schedule

Outline of the calendar for the work. This is necessarily rough and
subject to change. Usually include a Gantt chart

## Completed to date


## Six month plan
-->

{{< pagebreak >}}

# References

::: {refs}
:::
