---
title: Coping with climate change.
subtitle: Implications of the Yellow-Bellied Marmot's (*Marmota flavivientris*) body mass evolution in the last half-century.
author: Augustin Birot (300444988)
date: last-modified
affiliation: Department of Biology
comps-proposal: true # or false
committee:
  - name: Julien Martin
    role: supervisor 
  - name: Vincent Careau
    role: TAC member
  - name: Roslyn Dakin
    role: TAC member
degree: PhD


bibliography: [research_proposal.bib, packages.bib]
csl: american-statistical-association.csl

format:
  #html:
   bio-uo-proposal-pdf:
    toc: true
editor: 
  markdown: 
    wrap: 72
---


{{<pagebreak>}}

# Introduction

## Climate change (C.C.)

The importance of C.C. and its impact in the near future is no longer in doubt, *at the point where it's sad to have to remind people of them once again.*[@intergovernmentalpanelonclimatechangeipccClimateChange20222023]

* Broadly, C.C. is [...] (Polar melting, etc.)

* Which even impact human society (e.g., winter in Ottawa isn't the same anymore: Rideau Canal ice skating future is in jeopardy, the number of days with under -20°C is expected to severly decrease in the near future, etc. (!!! + FACT CHECK everything!))

* Main/Precise impacts of C.C. in natural environments

  * **Raising T°**: Explain + study case (!!!)

  * **Changing season length**: Explain + study case (try to find something at RMBL !!!)

  * **Environmental predictibility**: Explain + study case (!!!)

  * **Drought events**: Explain + study case (!!!)

  * **Extreme weather events**: Explain + study case (!!!)

  * etc.

Ok, so, now, how does C.C. and these precise perturbations impacts concretely natural population? *Study cases* (!!!)


## Body mass as a Life-History Traits (LHT)

Life history traits (LHT) are defined as phenotypic traits (i.e., individuals' characteristics) impacting directly the survival and reproduction (so the fitness) of an individual [@Roff1992TheEO].

[...]

For many species, body mass will indeed impact directly these two parameters

Energetic reserves => Impact individuals capacity; Honnest signal of good performance and ressources for sexual competition.

Not only an individual with a high body mass will have more energetic ressources, allowing it to be more resilient to environmental pressure (as this reserves can buffer a low ressource phase for example), but it can also indicate a better ability from the individual to gather ressources, showing better adaptation and survival skills in a given environment.
The first consequence of that being an increase of the individual survival with its body mass.
But also an indvidual with more energy will have better capacity to reproduce, and should be more appealing to potential mates in the case of sexual competition, therefore also increasing its reproduction success. Therefore, in many species, body mass can be indeed consider as a LHT. <!-- Okay, its very badly written but at least you have the beginning of something, add some citations and flesh out your reasoning -->

### Link with hibernation

The importance of body mass as a LHT is particularly true for a specific group of species: **Hibernating species**.
Indeed, to survive unfavorable season (usually winter in our latitudes) some species disperse to milder environment and other cope with it and find ways to survive through. One adaptation to that is hibernation <!-- FIND GENERAL LITERATURE ABOUT HIBERNATION!!! -->, which correspond to a state of torpor, individual keeping their metabolism at a survival minimum, surviving on their energetic ressources through the winter, emerging at the start of the favourable season. A prerequired adaptation for that kind of behaviour is therefore a large body mass, in order to have enough energetic reserve to stay without ressources for a full season. Even more, hibernating species are usually active only for a small part of the year, during which they must gather enough ressources, gain enough mass, to survive through next hibernation. So not only a prerequired adaptation is a big body mass, but also the capacity to gain quickly and efficiently body mass in a short amount of time, which represent a lot of challenges and therefore adaptation for the metabolism.

Furthermore, for some species, reproduction occurs right after the hibernation period, which means that before entering into hibernation, individuals much reach a high enough body mass not only to survive, but also to have enough energy left to reproduce right at the start of the active season. Hence, in that disposition, body mass is a keystone LHT for hibernating species biodemography.

<!-- {One of the most commonly known example of hibernating species in our society are the Marmots [@armitageMarmotBiologySociality2014]. [...]} (Is it a good strategy to bring the marmots (as a tribe, not the YBM yet) here, knowing that I wont talk about it till the end of the intro?) -->

In addition to that, we know that climate change is expected to have significant impact on LHT. An impact on body size has even been proposed to be a third universal response to climate change, alongside modification in phenology and geographic range [@gardnerDecliningBodySize2011; !!! *look for an article explaining the principles of the universal responses to global warming*]


### Expected effect of global warming on body mass

As reminded earlier, one of the most significant consequences of climate change is an increase in global temperature (which is why climate change is also commonly reffered to as *global warming*, although this term is often used as a rethoric by climate sceptics during cold winters and violent blizzards[^1]<!-- Okay it is poorly said but find a way to keep that in the final proposal -->).
This average temperature increase is suspected to influence phenotypic traits such as body mass or size. However, the direction of the response remains uncertain. Some authors argue that a shrinking body size could be one the universal C.C. response [@daufresneGlobalWarmingBenefits2009]. This hypotheses follows Bergmann's rules, which state that smaller body size should be expected in warmer environment as it raises the surface to volume ratio, thus favoring heat dissipation [@bergmanncRelationshipsHeatConservation1847].
However, as noted by @gardnerDecliningBodySize2011, a lack of large-scale comparative studies prevent us to demonstrate that this response could be universal. And indeed, several studies at higher latitude yield opposite results [i.e., increasing body mass, @guillemainWinteringFrenchMallard2010; @yom-tovRecentIncreaseBody2008; @sheridanShrinkingBodySize2011; @ozgulCoupledDynamicsBody2010]. At higher latitudes and altitude, climate change is synonym to milter conditions. As a consequence, individuals have access to a large food supply for a longer time and face less hard conditions during the less favorable season which overall is less energetically demanding, allowing individuals to become larger.

[^1]: "[...] Large parts of the Country are suffering from tremendous amounts of snow and near record setting cold. [...] Wouldn't be bad to have a little of that good old fashioned Global Warming right now!" Donald J. Trump, Jan 20, 2019.

<!-- IPCC 2022: "Evidence is weak for a consistent reduction in body size across taxonomic groups in terrestrial animals (low confidence) (Spiepielski et al. 2019)." -->


## Q.G. and animal models

Body mass and LHT shifts expected with climate change, **evolution** expected.

{To properly estimate the evolutionary signals of a phenotypic change, quantitative genetic gives us a method to decompose the total phenotypic variance ($V_P$) into it's genetic ($V_A$) and environmental ($V_{PE}$) part: $V_P = V_A + V_{PE}$.} <!-- Rephrase that, recheck the proper term and add citations, like Lynsh & Walsh -->
<!-- add a part about quantitative genetic in general <=> quantitative trait: a trait mediated by a lot of different genes, each playing a small part-->
Quantitative Genetics provides a well-established method for estimating the genetic component of an observed phenotype variation: the **Animal Models** [@kruukEstimatingGeneticParameters2004]. This method allows a robust estimation of the genetic variance in a trait affected by a large number of genes with small effects [i.e., a "quantitative trait", @kruukStudyQuantitativeGenetics2014] by fitting a mixed model with individual identity as a non-independant random effect, linked to a relatedness matrix between each individual, extracted from the population pedigree (i.e., parental link between the individual of a population, !!!).
This method has the advantage of being relatively simple to employ, allowing genetic parameter estimation directly from phenotypic data. Only parental links between individuals need to be known, making this method applicable to wild populations [@kruukEstimatingGeneticParameters2004; @Lynch1996GeneticsAA].
As emphasized by @kruukStudyQuantitativeGenetics2014, there is a pressing need for quantitative genetics studies on long-term wildlife populations as the most common problem in that kind of study is the lack of power, which could be adress with long-term studies. Such studies would improve our understanding of the relationship between animals and their environment and of the genotype-phenotype-environment relationship, especially in the context of global change.


<!-- [@10.1093/acprof:oso/9780199674237.001.0001] => Q.G. in the wild -->


## LHT coevolution

Traits can't evolve alone [@gouldSpandrelsSanMarco1979; @Roff1992TheEO] Gould & Lewontin (1979)

Need to show that with multivariate animal model, but no one has enough power for the models [@teplitskyEvolutionaryPotentialConstraints2014]

<!-- Résumé de ce chapitre, trouvable dans "Project Marmot I" (p. 105 - 110) -->

## POLS

[@dammhahnPacelifeSyndromesFramework2018]

## Phenotypic plasticity vs microevolution

Phenotype are expected to be the best fit for specific environment as a result of a long evolution by natural selection [i.e., individuals best adapted to their enviromnent will have better survival and reproductive success, @darwinOriginSpeciesMeans1859].
However, when this environment changes, as expected in today's context of climate change, individuals have two solution to avoid disapearance: **disperse** to a more favorable environment, or **adapt** to their new conditions via phenotypic change [@gienappEvolutionaryDynamicsResponse2014].
For adaptation, two further possibilities exist: **phenotypic plasticity**, defined as a change in phenotype expressed by a given genotype (!!! *probably Nussey or maybe Pigliucci, I think there's a book from the late 90' or early 00'*), which allows for a rapide response within an individual lifetime, is highly flexible and does not involve any changes at the genetic level; and **microevolution**, defined as a change in alleles frequencies in a population over time (!!! *needed?*).
{When an individual with a better-fitted phenotype for its new environment appears, it would have a better survival and more reproductive success. If this advantage relies on a heritable genetic difference (i.e., transmitted to its descendants, !!! *needed? something like Lynch & Walsh, for a definition of heritability sensu stricto*) the new genotype is going to rapidly increase in proportion in the population, ultimately replacing the old one} <!-- A bit long... -->.
Thus, this mecanisme can be slow but is a long-term solution when the ecological change is persistent. However if the change is transient, plasticity is a useful mechanism.
As noted by @dewittCostsLimitsPhenotypic1998 and @gardnerDecliningBodySize2011, phenotypic plasticity solely is unlikely to be the most optimal long-term response to climate change as it is usually a transient answer, presenting costs and limits [@dewittCostsLimitsPhenotypic1998], to a transient change. Furthermore, if the optimal response to the new environment is a canalized phenotype (i.e., very low phenotypic variance), plasticity can even be maladaptive [@nusseyEvolutionaryEcologyIndividual2007]. The expected optimal answer to a long-term environmental change, as caused by climate change, is evolution through natural selection.

Phenotypic pasticity and microevolution are thus not expected to be mutually exclusive. This is particularly evident in highly plastic traits such as body mass which can vary significantly up and down throughout an individual's life in response to among- and within-year changes in environmental conditions but can also change via microevolution at the population level over the same time period <!-- not the right moment? But make a good transition -->.

Nevertheless, as the consequences of theses mechanism can be highly different on the long term (evolution being more permanent than plasticity), quantifying the extent to which each of these mechanisms contributes to the observed change over a long study period remains a chalenging but fundamental task to understand the adaption and evolution of species. <!-- copy paste from body mass evolution manuscript, need a lot of re-writing (like the rest of this section) but need to keep the idea --> This is even more true today, as populations faces the numerous challenge brought by global climate change.

So Evolution and plasticity are not mutually exclusive, and even more, evolution can even have effect on plasticity itself.
*Transition with $I * E$ with the reaction norm framework [@nusseyEvolutionaryEcologyIndividual2007].*


## $\textbf{I * E}$, $\textbf{G * E}$ (individual variation in their plasticity)

Plasticity is usually studied using ***Reaction Norm framework*** (!!! *look at DeWitt & Scheiner, 2004; Schlichting & Pigliucci, 1998; West-Eberhard, 2003 from Thibaut's proposal intro*), so by studying the value of a phenotypic trait (e.g., body masse, size) in response to an environmental proxy (e.g., temperature, precipitation). A plastic response correspond to a different phenotypic value associated to a different environment, so by a slope different from 0 for the reaction norm [@nusseyEvolutionaryEcologyIndividual2007].
A reaction norm has two parameters: **Elevation** (or "Intercept" in statistical terms), which is the expected phenotypic value in the average environment; and the **Slope** (so "Slope" in statistical terms) corresponding to the linear regression of the phenotype over the environmental gradient, so to estimate the phenotypic response to the environment (i.e., phenotypi plasticity) <!-- Part extremely redundant for the sake of clarity -->.
So that's how phenotypic plasticity is usually studied. But, as emphasized by @nusseyEvolutionaryEcologyIndividual2007, a lot of information is loss if we just study that at the population level (i.e., by just doing a fixed linear model of the phenotype over the environment). First using a linear mixed model is crucial to estimate different elevations for each individual as comonly there's multiple observation per individual, thus not taking that into acount is violation of the independance application condition (cf every basic stat book <!-- do I need to add a citation here? -->).
So doing that allow us to estimate individual variance in the phenotype in the average environment and take into account the fact that all individual in the population doesn't have the same base phenotype. However, that kind of model doesn't allow to estimate individual-level difference in their plastic response to the environment, which can be really important biologically and so must be accounted for. @nusseyEvolutionaryEcologyIndividual2007 proposed a framework to estimate that, which is now fairly accepted and used: ***Random Regression***. Let's go back to the Linear Mixed Models (LMMs), the "classic" way is to add a random individual term on the intercept to estimate the "individual variance" by allowing the model to fit one intercept per individual, however in this compilation, the slope is the same for each individuals; now, if we add a random term also on the slope, not only the model estimate one intercept per individual, but also one slope (so basically, one reaction norm per individual), estimating thus the individual variance in the phenotype expected in the average environment (*elevation*) but also the individual variation in phenotypic variation (*slope*) which is commonly called $\textbf{I * E}$.

Once we've estimated this between individual variance in plasticity, we can you quantitative genetic methods (i.e., from Animal models, as explained before, but using Random Regression Animal Model: "RRAM", @nusseyEvolutionaryEcologyIndividual2007) to decompose this variance into it's genetic and permanent environment part to estimate the genetic variation in plasticity, which is called $\textbf{G * E}$. From that we can estimate the heritable variation in reaction norm's slope and have an idea of the evolutionary potential of phenotypic plasticity in itself.


So we have a really good method to do all that, however, as underlyned by @ramakersProbingVariationReaction2023, if the environmental proxy used for the reaction norm isn't good enough, a lot of this I * E is missed (what they called a "hidden I * E"). Furthermore, it's often impossible to identify the real driver of plasticity in natural conditions (Which is expected as natural environments are exceptionally complex systems, and individuals generally doesn't have only one environmental variable to deal with...), the real driver can be unknown, unmeasurable are a combination of a lot of different variables [@ramakersProbingVariationReaction2023]. In response to that, the authors proposed a new method, using the Environment Specific Mean phenotype (ESM). Although they've shown this method is indeed efficient, they emphasized that we still need a really good knowledge of the studied system, and really specific conditions to be more efficient that the "regular" method. Hence, although it's a progress, more work is needed to deal with this I * E detection in natural environment problem.

The good news is that we have a really promising, fairly new statistical method that could be applied to that: the "Double Hierarchical Generalized Linear Model" (DHGLM, !!! *find citation + maybe explain a little the method*) [...] from this method we can estimate the *variance in the residual variance* (i.e., "$\boldsymbol{V_{V_e}}$" by estimating the residual variance attributable to each individual). Knowing what these models does (*When I'll write it*), we're confident that it can be applied to developp a more powerfull method to detect individual variation in plasticity (I * E) in natural systems. Indeed, a $V_{V_e}$ significantly greater than 0 should indicate that there is some I * E in the system and would justify investigation by testing different environmental proxies. In contrast, except in very specific conditons, almost impossible in natural conditions (i.e., **perfect** contrary effect), a $V_{V_e}$ equal to 0 indicate that there's no individual variation in plasticity in the studied system. From this method we should be a way to free ourselves from the problems of bad proxies. And obviously, as for the random regreession with the RRAM, we can adapt this method we quantitative genetics to also detect $G * E$ if $I * E$ is detected.

This new method is thus very promising, however, as most new statistical methods, it often poorly understood/used. Hence, a general framework (supervised by someone with good statistical knowledge, i.e., Julien) including the application on how DHGLM could (should) be used to detect I * E would be really helpful for the comunity (especially with the rising need to study efficiently phenotypic plasticity and it's evolution, in natural condition with climate change).


Link with body mass, individual can vary in their growing speed <=> Reaction norm/Plasticity change over time => Evolution directly on the plasticity <=> individual answer to the condition change would be increase their response (i.e., body mass increase within the active season). It would make sense with bet-hedging framework for example


### Bet-hedging

(!!! *Ref about bet-hedging*)

Bet on the best fitness for the long term (even if it can mean lower a bit your immediate fitness) to cope with an unpredictable environment.
With climate change, environment are less predictable than ever (maybe even more in alpine habitat? !!! *fact check + ref about envmt predictability + focus on alpine habitats*).
Thus bet-hedging, for example increase you body size to "buffer" is not a crazy strategy, but can be risky in the future => potential phenological mismatch (i.e., !!! *def + ref about what phenological mismatch is*), bet-hedging is a bet, so you're not sure to win in the end, and it can end in maladaption...



## Species and study site

A wild Yellow-Bellied Marmot (*Marmota flaviventris*, "YBM") population in the Upper East River Valley, Colorado, USA, is the subject of one of the longest-term study in the world (1962 - today).
YBM is a ground-dwelling sciurid (rodentia, sciurideae) inhabiting alpine habitats in western North America with a life cycle divided between an "active season" representing approximately a third of the year (from May to September) where individuals must forage to reach a threshold body mass in order to survive hibernation for the remainder of the time [@armitageMarmotBiologySociality2014].
Individuals experience high seasonal flutuation in body mass, with a critical threshold to be reached before the onset of hibernation in order to 1) survive through the next active season and 2) have sufficient energy left for hibernation [which occurs in the first weeks of the active season, @armitageVernalBehaviourYellowbellied1965; @armitageMarmotBiologySociality2014].
Consequently, body mass is consider being a critical LHT for the marmots.
YBM lives in colonies composed usually by one or more matriline with on adult males, multiple adult females and their offspring [@armitageMarmotBiologySociality2014]. Our population is composed of seven main colonies divided between an "up" and a "down valley" with a elevation difference arround 300m ("up" = 3,000m; "down" = 2,700m) implying some difference in weather [e.g., delayed snowmel and vegetation growth onset, temperature difference up to 2 °C, @blumsteinYellowbelliedMarmotsMarmota2004; @armitageMarmotBiologySociality2014] and so delayed emergence up to two weeks in the up-valley [@blumsteinSOCIALEFFECTSEMERGENCE2009; @monclusYellowbelliedMarmotsNot2014].
This two differents condition offers an amazing opportunity to test the impact on environment on several factors while working in natural conditions.

This hibernation (life) cycle is highly environemntally dependant, with the onset and end of the active season believed to be mediated mostly by weather variable such as temperature and snow cover of the region [@armitageMarmotBiologySociality2014]. Thus, body mass is expected to be a keystone phenotypic trait for the marmots. It is therefore crucial to understand how this trait and this species responds to global warming, both for conservaion purposes and to elucidate links between phenotype and environment.



### Body mass increase in YBM

An important body mass increase has been observed in this population over the past half-century (estimated arround 600 g for the adult females). Precedent studies attributed this major change mostly to phenotypic plasticity [@ozgulCoupledDynamicsBody2010]. This hypothese made in fact a lot of sense, with climate change active season is getting longer (milder condition, higher temperature, less snow, shorter winter, etc.), hence marmots have more time to forage, gain weight, and the hibernation period is getting shorter so less time for the individuals to lose mass, at the end of the day, we have heavier individuals, makes sense!
However, using animal models to properly assess the genetic attributable part of this change, estimating explicitely the body mass' evolutionary signal for the adult females over the time cohort (i.e., year of birth) during the study period, we found an increase, at the genetic scale, estimated arround 400 g, with a heritability of $56\%$ (Birot & Martin, Manuscript in progress). So, in fact, arround two third of the body mass increase seems to be due to evolution, not just plasticity.
Furthermore, although the lengthening active season is indeed a good potential explanation for the body mass increase through phenotypic plasticity, it doesn't match with the observed evolutionary signal. If the main selective pressure on body mass is survival through hibernation (i.e., heavier individuals having more chance to survive through winter as they have more ressources), then the expected evolutionary response (i.e., average body mass increase) is occuring when the pressure is decreasing, which doesn't makes sense!

{{<pagebreak>}}

## Research objectives & Methodology

Considering that this major phenotypic shift for the YBM in the last half-century has been miss considered, knowing that there is in fact a strong evolutionary signal, we need to reconsider the evolutionary scenario behind this body mass increase.

I will explore which environmental factors could have triggered this shift, but also the mechanism behind this increase and finally the potential implication for the population's future. <!-- DEVELOPP, but not too much as you developp in the "chapters" -->



<!--
# Research objectives and hypotheses

Specific hypotheses (in the case of hypothesis-testing science) or
objectives (in the case of descriptive science), and the underlying
logic, for each proposed thesis chapter (minimum three for PhD students)

# Proposed methodologies

Include outlines of data collection and data analysis

Outline any special skills, equipment, or techniques that will be
required.

In case you want to add a wide table that needs to be on a landscape
page you can do like in @tbl-sideway. Have a look at the `qmd` for the
code.

You need to put the table in a pandoc div (delimited by :::) indicating
the side way rotation

It also works for figures just use `sidewaysfigure` instead of
`sidewaystable`

::: {.sidewaystable data-latex=""}
```{r}
#| label: tbl-sideway
#| echo: false
#| tbl-cap: This is a wide table
library(gt)
exibble |>
  gt(
    rowname_col = "row",
    groupname_col = "group",
    locale = "fr"
  ) |>
  fmt_number() |>
  fmt_date(
    columns = date,
    date_style = "yMEd"
  ) |>
  fmt_datetime(
    columns = datetime,
    format = "EEEE, MMMM d, y",
    locale = "en"
  ) |>
  tab_options(table.width = pct(100))

```
:::

-->


{{<pagebreak>}}


# Chapter 1 - Mechanisms
*Marmot's Biology: What mechanisms are behind the body mass increase?*

As emphasized in the introduction, we know that the body mass has increased in the last half-century, however we still know very few about it. And for starter, we don't know yet what has changed in the marmots in the last decades: do they have a bigger constitution (i.e., are they born bigger, their Intercept in statistical term), or does their growing capacity has increased through time (in statistical term again, this would correspond to their slope)?

### $\textbf{I * A}$ and $\textbf{G * A}$
A: AGE => Reaction norm over individual lifetime rather than Environmental gradient

Growth? Baseline? Both?


Double random (Intercept, Slope)


```{r}
#| echo: false
#| message: false
#| warning: false
#| eval: true
#| label: fig-juvenile-trend
#| fig-cap: Body mass trend over time cohort for females (black) and males (red) juveniles compared between a) the beginning of the active season (birth weight) and b) the end of the season (mass on August 15th).

source("figures/juvenile_mass.r")

cowplot::plot_grid(plot2_jun, plot2_aug, ncol = 2)
```

{{<pagebreak>}}

# Chapter 2 - Methodology
*Methodology: $I * E$ detection with double random mixed models*

[@nusseyEvolutionaryEcologyIndividual2007] --> double random

So we're doing something different --> examining the residuals of the model (if I * E, still a lot of residual variance ?)


**Look at this one: [@westneatBiologyHiddenResidual2015a]


DHGLM, brms, Julien's code

Vve (Variance dans la variance résiduel, estime la variance résiduel pour chaque individu et regarde la variance dans cette variance résiduelle, si I * E Vve > 0)

Attention aux modèle débalancés si pas d'effet fixes
corrige pour les variations par effet fixes, puis test pour le I * E, si y'en a tu pexu chercher la variable environnemental pour lesquels on a de la variation dans la plasitcité (I * E)

Ned Dochtermann


<!-- Okay, so now you know what is the problem about estimating/detecting I * E, you know what is a DHGLM (cause I've explained everything in the intro) now here's what I'm gonna do

Simulations to see how we can estimate it (data with and without I * E => fit the models, check how the different parameters reacts and how we interpret it, and also how we can present them)

Simulation to test the different bias (Sampling bias, etc.) -->


{{<pagebreak>}}

# Chapter 3 - Triggers
*Marmot's Biology: Which environemntal factors have triggered the phenotypic shift?*

E1 - E10 (T°, Precipitation, ...), Seasonal Gradient

Predators, Diet?

{{<pagebreak>}}

# Chapter 4 - Implications
*Marmot's Biology: What could be the implications of that for the population's future?*

Manuscript models Body Mass/active season with survival
=> Phenological mismatch?? (e.g., thermal stress)


{{<pagebreak>}}

# Significance and impacts



<!--
# Expected Products

A list of the (usually 1/3) articles expected for completion of the
MSC/PhD. Think of each as a fantasy abstract.

1.  The Evolution of Human Uniqueness. [Bob Hette & Julien
    Martin.]{.underline}

    There is wide agreement that humans are unique. But it is rarely
    asked: Have they always been unique? In this paper, we derive a
    general uniqueness theorem that states that any unique species must
    have been less unique at some point in the past. This theorem has
    implications for the achievement of world peace.

2.  Niche Construction, Disease Transmission, and Global Warming: Chaos
    Deconstructs the Block Chain. [Bob Hette, Igor Tex & Julien
    Martin]{.underline}

    How many buzz words can be combined in a title? In this paper, we
    find that the natural limit on buzz words in any specific scholarly
    field is always given by the triangular number $T_n$ where $n$ is
    one more than the number of tenured faculty in the field.

3.  A Bayesian Hierarchical Survival Analysis of Counting Sheep. [Bob
    Hette & Julien Martin]{.underline}

    Just like all those other models of counting sheep, but this one
    will be Bayesian.

# Scientific value and rationale for the research

# Course selection

not necessarily needed for comprehensive exam proposal

# Schedule

Outline of the calendar for the work. This is necessarily rough and
subject to change. Usually include a Gantt chart

## Completed to date


## Six month plan
-->

{{< pagebreak >}}

# References

::: {refs}
:::
